Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-11-01 pkg/disttree/inst/draft/draft.Rnw
small touch-ups in abstract/intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2372 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4eb56ab7907fb1c328e898785cfe442533ec4c10

\Abstract{
To obtain a probabilistic model for a dependent variable based on some set of
explanatory variables, a distributional approach is often adopted where the
parameters of the distribution are linked to regressors. In many classical
models this only captures the location of the distribution but
over the last decade there has been increasing interest in distributional
regression approaches modeling all parameters including location, scale, and
shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
mean and variance of a Gaussian response and is particularly
popular in weather forecasting. 
% More generally, the GAMLSS framework allows to establish generalized additive 
% models for location, scale, and shape with smooth linear or nonlinear effects.
\hlc{Less specifically, generalized additive models for location, scale, and shape (GAMLSS) provide a framework where each distribution parameter is modeled separately capturing smooth linear or nonlinear effects.}
%
However, when variable selection is required and/or there are
non-smooth dependencies or interactions (especially unknown or of high-order),
it is challenging to establish a good GAMLSS. A natural alternative in these
situations would be the application of regression trees or random forests but,
so far, no general distributional framework is available for these. Therefore,
a framework for distributional regression trees and forests is proposed that blends
regression trees and random forests with classical distributions from the GAMLSS
framework as well as their censored or truncated counterparts.
%
To illustrate these novel approaches in practice, they are employed to obtain
probabilistic precipitation forecasts at numerous sites in a mountainous
region (Tyrol, Austria) based on a large number of numerical weather
prediction quantities. It is shown that the novel distributional regression forests
automatically select variables and interactions, performing on par or often even
better than GAMLSS specified either through prior meteorological knowledge or a
computationally more demanding boosting approach.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}

@@

\hlc{However, distributional forests should not be considered as a 
replacement of GAMLSS but rather as a complementing method in the field of
distributional models, similar to the way how GAM and random forests are complements 
in the conditional mean regression toolbox.}

In the following, particular focus is given to distributional forests as a method for
obtaining probabilistic forecasts by leveraging the strengths of random forests: the 
ability to capture both smooth and abruptly changing functions along with simultaneous selection of variables and possibly complex interactions. Thus, these properties make the method particularly appealing in case of many covariates with unknown effects and 
interactions where it would be challenging to specify a distributional regression model 
like GAMLSS.

@@

In weather forecasting, these properties are especially appealing in 
mountainous regions and complex terrain where a wide range of local-scale
effects are not yet resolved by the NWP models. 
Thus, effects with abrupt changes and possibly nonlinear interactions 
might be required to account for site-specific unresolved features.
To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
More specifically, a Gaussian distribution left-censored at zero, is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of observing no precipitation 
on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
established based on data from the same month over the years 1985--2012 including
80~covariates derived from a wide range of different NWP quantities. As
Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
distributional forecast for each day -- here for one specific day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as learning data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{sec:precipitation} we investigate the
performance of distributional forests in this forecasting task in more detail.
Compared to three alternative zero-censored Gaussian models distributional forests
perform at least on par and sometimes clearly better while requiring no meteorological
knowledge about the atmospheric processes which drive formation of precipitation
for the model specification. The three alternatives are: a standard ensemble model
output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
80~available regressors.

>>>>

\Abstract{
To obtain a probabilistic model for a dependent variable based on some set of
explanatory variables, a distributional approach is often adopted where the
parameters of the distribution are linked to regressors. In many classical
models this only captures the location of the distribution but
over the last decade there has been increasing interest in distributional
regression approaches modeling all parameters including location, scale, and
shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
mean and variance of a Gaussian response and is particularly
popular in weather forecasting. 
\hlc{Moreover, generalized additive models for location, scale, and shape (GAMLSS)
provide a framework where each distribution parameter is modeled separately
capturing smooth linear or nonlinear effects.}
%
However, when variable selection is required and/or there are
non-smooth dependencies or interactions (especially unknown or of high-order),
it is challenging to establish a good GAMLSS. A natural alternative in these
situations would be the application of regression trees or random forests but,
so far, no general distributional framework is available for these. Therefore,
a framework for distributional regression trees and forests is proposed that blends
regression trees and random forests with classical distributions from the GAMLSS
framework as well as their censored or truncated counterparts.
%
To illustrate these novel approaches in practice, they are employed to obtain
probabilistic precipitation forecasts at numerous sites in a mountainous
region (Tyrol, Austria) based on a large number of numerical weather
prediction quantities. It is shown that the novel distributional regression forests
automatically select variables and interactions, performing on par or often even
better than GAMLSS specified either through prior meteorological knowledge or a
computationally more demanding boosting approach.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}

@@

In the following, particular focus is given to distributional forests as a
method for obtaining probabilistic forecasts by leveraging the strengths of
random forests: the  ability to capture both smooth and abruptly changing
functions along with simultaneous selection of variables and possibly complex
interactions. Thus, these properties make the method particularly appealing in
case of many covariates with unknown effects and  interactions where it would be
challenging to specify a distributional regression model  like GAMLSS.
%
\hlc{However, distributional forests should not be considered as a replacement
of GAMLSS but rather as a complementing technique for flexible distributional
regression -- much like GAMs and random forests are complements for conditional
mean regression.}

@@

In weather forecasting, the flexibility of distributional forests is especially appealing in 
mountainous regions and complex terrain where a wide range of local-scale
effects are not yet resolved by the NWP models. 
Thus, effects with abrupt changes and possibly nonlinear interactions 
might be required to account for site-specific unresolved features.
To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
More specifically, a Gaussian distribution left-censored at zero, is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of observing no precipitation 
on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
established based on data from the same month over the years 1985--2012 including
80~covariates derived from a wide range of different NWP quantities. As
Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
distributional forecast for each day -- here for one specific day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as learning data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{sec:precipitation} we investigate the
performance of distributional forests in this forecasting task in more detail.
Compared to three alternative zero-censored Gaussian models distributional forests
perform at least on par and sometimes clearly better while requiring no meteorological
knowledge about the atmospheric processes which drive formation of precipitation
for the model specification. The three alternatives are: a standard ensemble model
output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
80~available regressors.
